---
title: "explore_species"
author: "Aurore A. Maureaud"
date: '2022'
output:
  html_document: 
    toc: true
    theme: flatly
  word_document: default
  pdf_document: default
---

Exploring the flags and processing methods on several crab species, here:

```{r species, echo=FALSE}
spp <- "Dyspanopeus texanus"
print(spp)
```

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(sf)
library(tidyverse)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)
world <- ne_countries(scale = "medium", returnclass = "sf")
library(egg)
library(raster)
sf::sf_use_s2(FALSE)
library(RColorBrewer)
library(units)
library(rfishbase)
library(readxl)

# load occ data for crabs
dat_occ_nf <- get(load("~/Yale University/MOL-inverts/TaxToSDM/outputs_MOL/crabs/DAT_OCC_CRABS_9JUN2021.RData"))

# load bpow layer
bpow <- st_read("~/Yale University/Marine Biogeography/bpow/outputs/bpow/bpow_p8s5_abyssal_corr.shp")

# identify duplicated data via sources
# even with this procedure, 21 duplicates remain because of different catalog numbers
# don't know if they then should be considered separate or not
dups <- dat_occ_crabs %>% 
  filter(duplicated(occ_id)) %>% 
  dplyr::select(occ_id) %>% pull()

dat_occ_crabs <- dat_occ_crabs %>% 
  filter(!(occ_id %in% dups & source == "OBIS"),
         !(occ_id %in% dups & source == "GBIF")) %>% 
  distinct()
rm(dups)

# fao fishing areas
fao <- st_read("~/Yale University/Marine Biogeography/regions_base/fao/World_Fao_Zones.shp")

# load traits
traits <- read.csv("~/Yale University/MOL-inverts/TaxToSDM/outputs_MOL/crabs/traits/master_traits_22JUL2021.csv")
high_traits <- read.csv("~/Yale University/MOL-inverts/TaxToSDM/outputs_MOL/crabs/traits/high_traits_22JUL2021.csv")

# FAO areas and species
fao_slb <- read_excel("~/Yale University/Crabs/SLB_FAO.xlsx")

```

## 1. No filter

Occurrence points based on the raw occurrence data from OBIS and GBIF, with no filtering, but after the OBIS-GBIF deduplication methods applied.

```{r step 1, echo = FALSE}

  dat_spp <- dat_occ_crabs %>% 
    filter(accepted_name == spp)
 
  st_as_sf(dat_spp, coords = c("longitude","latitude"), crs = st_crs(world)) %>% 
    ggplot() + geom_sf(shape = 3) +
    geom_sf(data = world, fill = "brown", alpha = 0.4, color = NA) +
    theme_bw() +
    ggtitle(paste0(length(unique(dat_spp$occ_id)), " points, ", spp))
```

## 2. Occurrence filters

Occurrence spatial points are filtered based on flags of sveeral categories:

- missing year
- problem with coordinate (0 island or missing)
- removing the points with high coordinates uncertainty
- remove on land points

```{r step 2, echo = FALSE}

  dat_spp_2 <- dat_spp %>% 
    filter(is.na(flag_year),
         is.na(flag_coordinate),
         flag_coord_uncertainty %in% c("NA","missing_uncertainty"),
         #is.na(flag_outlier),
         is.na(flag_realm)
         )
 
  st_as_sf(dat_spp_2, coords = c("longitude","latitude"), crs = st_crs(world)) %>% 
    ggplot() + geom_sf(shape = 3) +
    geom_sf(data = world, fill = "brown", alpha = 0.4, color = NA) +
    theme_bw() +
    ggtitle(paste0(length(unique(dat_spp_2$occ_id)), " points, ", spp))
  
```

## 3. Habitat information and BPOW

The depth range expected of the species is used to check the bathymetric groups in which the species is found:

- we used the depth range of the species from SeaLifeBase species profiles
- we translated that depth range into broad bathymetric categories used in the BPOW biogeographic regionalization layer:

1. 0-800m: coastal and upper bathyal
2. 800-3,500m: bathyal
3. 3,500-6,500m: abyssal
4. 6,500m: hadal trenches

- we compared to the two

```{r habitat, echo=FALSE}
i <- which(traits$canonical == spp)

# deduct bathymetric group(s)
groups_spp <- c("coastal", "bathyal","abyssal","hadal")
if(traits$depth_range_shallow[i]>800){groups_spp[1] <- NA}
if(traits$depth_range_deep[i]<6500){groups_spp[4] <- NA}
if(traits$depth_range_deep[i]<3500){groups_spp[3] <- NA} 
if(traits$depth_range_deep[i]<800){groups_spp[2] <- NA}
groups_spp <- groups_spp[!is.na(groups_spp)]
print(paste0("depth range: ", traits$depth_range_shallow[i],"-",traits$depth_range_deep[i]))
print(paste0("bathymetric groups: ", groups_spp))

```

```{r bpow, echo=FALSE}

dat_spp_3 <- st_as_sf(dat_spp_2, coords = c("longitude","latitude"), crs = st_crs(world))
bpow_spp <- st_join(dat_spp_3, bpow, left=TRUE, largest=FALSE) %>% 
  st_drop_geometry() %>% 
  dplyr::select(ID) %>% 
  distinct() %>% 
  pull()

bpow <- bpow %>% 
  mutate(occ = ifelse(ID %in% bpow_spp, TRUE, FALSE),
         slb = ifelse(type %in% groups_spp, TRUE, FALSE),
         slb_occ = ifelse(slb ==TRUE & occ == TRUE, "SLB & Occurrences",
                          NA_character_),
         slb_occ = ifelse(slb=="TRUE" & occ=="FALSE", "SLB", slb_occ),
         slb_occ = ifelse(slb=="FALSE" & occ=="TRUE", "Occurrences", slb_occ))

bpow %>% filter(!is.na(slb_occ)) %>% ggplot() + geom_sf(aes(fill = slb_occ), alpha= 0.4) +
  scale_fill_manual(values = c("forestgreen","blue","yellow"),
                    breaks = c("SLB & Occurrences","SLB","Occurrences")) +
  geom_sf(data = world, fill = "brown", alpha = 0.4) +
  geom_sf(data = dat_spp_3, shape =3) +
  ggtitle("Observation versus Expert knowledge for BPOW")

```

## 4. FAO zone presence

The fishing FAO areas are used to incidate the expected presence of the species in several sub-ocean areas:

- the reference used is the list of FAO areas from SeaLifeBase species profiles
- the FAO shapefile is used to indicate the FAO areas in which the species is found based on the occurrence spatial points
- the two are compared

```{r step fao, echo=FALSE, warning=FALSE}
fao_spp <- fao_slb %>% 
  mutate(canonical = paste(Genus, Species, sep=" ")) %>% 
  filter(canonical == spp) %>% 
  dplyr::select(AreaCode) %>% 
  pull()

dat_spp_fao <- st_join(dat_spp_3, fao, left=TRUE, largest=FALSE)
dat_spp_fao <- unique(dat_spp_fao$zone)

fao <- fao %>% 
  mutate(slb = ifelse(zone %in% fao_spp, TRUE, FALSE),
         occ = ifelse(zone %in% dat_spp_fao, TRUE, FALSE),
         slb_occ = ifelse(slb=="TRUE" & occ =="TRUE", "SLB & Occurrences", NA_character_),
         slb_occ = ifelse(slb=="TRUE" & occ=="FALSE", "SLB", slb_occ),
         slb_occ = ifelse(slb=="FALSE" & occ=="TRUE", "Occurrences", slb_occ))

fao %>% filter(!is.na(slb_occ)) %>% ggplot() + geom_sf(aes(fill = slb_occ), alpha= 0.4) +
  scale_fill_manual(values = c("forestgreen","blue","yellow"),
                    breaks = c("SLB & Occurrences","SLB","Occurrences")) +
  geom_sf(data = world, fill = "brown", alpha = 0.4) +
  geom_sf(data = dat_spp_3, shape =3) +
  ggtitle("Observation versus Expert knowledge for FAO zones")

```

